\thispagestyle{empty}

%----------HEADING-----------------
\begin{tabular*}{\textwidth}{l@{\extracolsep{\fill}}r}
  \textbf{\href{http://nmorse.com/}{\Large Nathan Morse}} & \href{mailto:nam@psu.edu}{nam@psu.edu}\\
  Curriculum Vitae & \href{http://nmorse.com/}{www.nmorse.com} \\
\end{tabular*}


%-----------EDUCATION-----------------
\section{Education}
  \resumeSubHeadingListStart
    \resumeEducation
      {Pennsylvania State University}{University Park, PA}
      {PhD in Political Science (major: olitical methodology)}{Expected: May 2024}
      {MA in Political Science (major: American politics)}{May 2022}
    \resumeSubheading
      {University of Florida}{Gainesville, FL}
      {BA in Statistics and Political Science}{May 2019}
  \resumeSubHeadingListEnd
  

%-----------EXPERIENCE-----------------
\section{Experience}
  \resumeSubHeadingListStart
    \resumeSubheading
      {Pennsylvania State University}{University Park, PA}
      {Department of Political Science}{Aug 2019 - Present}
      \resumeItemListStart
        \resumeItem{Methods Preceptor (Fall 2023 - Present)}
          {Assistant instructor for graduate-level methods seminars. Involves guiding students on assignments, providing feedback on work, and grading.}
        \resumeItem{Instructor (Spring 2023)}
          {Taught Introduction to Political Research, an undergraduate course covering research design, introductory statistics, and AI tools for political science.}
        \resumeItem{Teaching Assistant (Fall 2020 and Fall 2022)}
          {Tutored students in R coding, helped students with questions, and graded assignments.}
        \resumeItem{Research Assistant (all other semesters)}
          {Collected, visualized, and analyzed data for research projects on topics ranging from media coverage of police protests to public perceptions of economic inequality (for Ray Block Jr., Suzanna Linn, Christopher Zorn, and Eric Plutzer). }
      \resumeItemListEnd
    \resumeSubheading
      {University of Florida}{Gainesville, FL}
      {Department of Political Science}{Jan 2019 - Aug 2019}
      \resumeItemListStart
        \resumeItem{Research Assistant}
          {Compiled voter registration files from around the country and contributed code to an open-source framework for standardizing state voter files (for Michael McDonald).}
      \resumeItemListEnd
        \resumeSubSubheading
         {J. Wayne Reitz Union}{Apr. 2017 - May 2019}
         \resumeItemListStart
            \resumeItem{Associate for Assessment}
              {Data scientist for student affairs. Fielded surveys, managed databases, analyzed data, designed reports, and presented findings for more than 10 departments and programs.}
         \resumeItemListEnd
        \resumeSubHeadingListEnd